\ttitle

### 2015-06-06 Chapters/Introduction.tex
Changed todo to easy-todo package. Works
Added Todos in introduction
Added a whole pile of Todos
- Time to take a break
- There is a pile of stuff to do in introduction still
- AAaaaaaaaaaaaaaaaaaarggggggggggggggggghhhhhhhhhhhhhh

\section{Comorbidity and Risk Stratification}
\label{sec:comorbidity_risk_stratification}
\subsection{Comorbidity}
Comorbidity is defined as the presence of or the effect of other diseases that a patient has in addition to the primary disease of interest. The presence of comorbid conditions is associated with adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010} and often limits therapeutic options available due to the associated complications or side effects of surgery or chemoradiotherapy.\parencite{sandroussi_sociodemographics_2010}

@@

\subsection{Risk Stratification}
Physiological fitness or reserve may be defined as the ability of the patient’s organ systems to respond appropriately and adequately to the stress of major surgery. Major surgery places a significant physiological stress on multiple organ systems, especially the cardiorespiratory system. The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and the postoperative recovery plays a major role in determining short-term outcomes.

@@

\subsection{Static Versus Dynamic Testing}

Objective measurement of oxygen delivery at the tissue level at times of physiological stress allows for identification of patients who may struggle during the perioperative phase. Identification of such high-risk patients allows not only for improved patient selection, but also for risk-stratified, anaesthetic and postoperative critical care. Preoperative risk stratification will also allow for prehabilitation of these patients in an attempt to improve outcomes.

@@

\section{Cardiopulmonary Exercise Testing}
%History of CPET
\subsection{History of CPET in Surgery}

@@

\subsubsection{Ventilatory equivalents method}
[...]

@@

\subsection{Breathing frequency, $B_f$}

\section{Role of CPET in preoperative assessment}
\subsection{General Surgery}
Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery.

Older and co-workers conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery. All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring. The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg. There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes. They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55) while only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, $p<0.001$). Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$).\parencite{older_preoperative_1993} 
In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise. High risk patients ($\dot{V}_{O_2}AT<11$ml/kg/min, n=153) were admitted to intensive care after surgery and had a mortality due to cardiovascular causes of 4.6\%. Moderate risk group ($\dot{V}_{O_2}AT>11$ml/kg/min and either $\dot{V}_E/\dot{V}_{CO_2}>35$ or evidence of myocardial ischaemia, n=115) was managed in the high dependency unit after surgery and had a cardiovascular mortality of 1.7\%. Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality.\parencite{older_cardiopulmonary_1999} Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient. \parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}

@@

\subsection{Relationship between systemic inflammation and comorbidity}

@@

\subsection{Impact of jaundice on the immune system}
%nehez_compromise_2002 - get other citations from this one

@@

\section{Body Composition}
Make this a brief description of body composition.

>>>>

\section{Comorbidity and Risk Stratification}
\label{sec:comorbidity_risk_stratification}
\todo{Comorbidity and Risk Stratification - needs major tidying up - will take several days}
\subsection{Comorbidity}
Comorbidity is defined as the presence of or the effect of other diseases that a patient has in addition to the primary disease of interest. The presence of comorbid conditions is associated with adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010} and often limits therapeutic options available due to the associated complications or side effects of surgery or chemoradiotherapy.\parencite{sandroussi_sociodemographics_2010}

@@

\subsection{Risk Stratification}
\todo{Intro - risk stratification}
%Include other measures - discuss POSSUM, APACHE, mGPS, EPASS and everything else here.
Physiological fitness or reserve may be defined as the ability of the patient’s organ systems to respond appropriately and adequately to the stress of major surgery. Major surgery places a significant physiological stress on multiple organ systems, especially the cardiorespiratory system. The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and the postoperative recovery plays a major role in determining short-term outcomes.

@@

\subsection{Static Versus Dynamic Testing}
\todo{Static Versus Dynamic Testing}
Objective measurement of oxygen delivery at the tissue level at times of physiological stress allows for identification of patients who may struggle during the perioperative phase. Identification of such high-risk patients allows not only for improved patient selection, but also for risk-stratified, anaesthetic and postoperative critical care. Preoperative risk stratification will also allow for prehabilitation of these patients in an attempt to improve outcomes.

@@

\section{Cardiopulmonary Exercise Testing}
\todo{A brief history of CPET in general}
\todo{The physiological basis of CPET}

@@

\subsubsection{Ventilatory equivalents method}
\todo{Ventilatory equivalents method}

@@

\subsection{Breathing frequency, $B_f$}
In healthy individuals, breathing frequency is one of the most important respiratory responses to exercise in order to increase oxygen uptake. However, a variety of respiratory disorders including obstructive disorders such as chronic obstructive pulmonary disease or restrictive disorders such as interstitial lung disease can impair this response. Morbidly obese patients may also have additional restrictions on respiratory response due to their body habitus. On the other hand, hyperventilation in an otherwise healthy patient may make interpretation of the results difficult although the analysis of trends in other parameters such as $P_{ET_{O_2}}$ and $P{ET_{CO_2}}$ aids in the diagnosis of hyperventilation versus true respiratory problems.

\section{Role of CPET in preoperative assessment}
\label{sec:cpx_roleInAssessment}
\subsection{General Surgery}

Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery. Older and co-workers conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery. All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring. The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg. There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes. They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55) while only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, $p<0.001$). Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$).\parencite{older_preoperative_1993}

In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise. High risk patients ($\dot{V}_{O_2}AT<11$ml/kg/min, n=153) were admitted to intensive care after surgery and had a mortality due to cardiovascular causes of 4.6\%. Moderate risk group ($\dot{V}_{O_2}AT>11$ml/kg/min and either $\dot{V}_E/\dot{V}_{CO_2}>35$ or evidence of myocardial ischaemia, n=115) was managed in the high dependency unit after surgery and had a cardiovascular mortality of 1.7\%. Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality.\parencite{older_cardiopulmonary_1999} Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient. \parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}

@@

\subsection{Relationship between systemic inflammation and comorbidity}
\todo{Relationship between systemic inflammation and comorbidity}

@@

\subsection{Impact of jaundice on the immune system}
%nehez_compromise_2002 - get other citations from this one
\todo{Impact of jaundice on the immune system}

@@

\section{Body Composition}
\todo{Introduction - body composition}
